A Classical Density-Functional Theory for Describing Water Interfaces

### 2011-12-02 paper/paper.tex
Work on paper, combining intro, deleting unnecessary figures and sections, cleaning up 1 rod section. Tested everything.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
{We develop a classical density functional theory for water by combining the
fundamental-measure theory functional--which describes hard sphere fluids--with
attractive interactions based on the Statistical Associating Fluid Theory
(SAFT-VR). Using this approximation with an additional length scale parameter, we are able to describe the vapor pressure, saturated liquid density, surface tension, and enthalpy of water, as well as the number of hydrogen bonds and density oscillations formed near hydrophobic rods and spheres immersed in water. 
\needsworknow{For two rods being pulled apart, a transition occurs 
from a vapor state to a liquid state in the area between
the rods at approximately $(\pi-2)$ multipled by the radius. Even 
for nanoscale rods (radius 0.5~-~1.0~nm), we find that in the
vapor state, the force can be approximated
classically and is equal to 144~mN/m, or twice the bulk surface tension.
(Keep this? Add spheres stuff?)}
This functional is computationally efficient and provides a foundation for 
future work describing interactions in aqueous solution on an
atomic level.}
\end{abstract}
\maketitle

{\begin{itemize}
\color{blue}
\item Put finished things up here
\item[JRH] Rewrite abstract (copying from project)
\color{red}
\item[JRH] Rewrite and join introductions
\item[JRH] Work on abstract
\item[DJR] Read and edit changes to abstract
\item[DJR] Look through theory section for stuff to cut
\item[DJR,JRH] Find figures to cut
\item[JRH] Cut the test-particle section
\item[JRH] Decide what to do with slit (cut?)
\item[DJR] Write text (or clean up existing) for hard sphere solutes
\item[EJK] Add 4-rods figures
\item[EJK] Add 4-rods text
\end{itemize}
}

@@

Due to its large dielectric constant, small molecular size and strong
hydrogen-bonding network, water as ``the universal solvent'' are as
interesting as its anomalous thermodynamic properties.

\begin{figure}
\includegraphics[width=6cm, angle=270]{figs/correlation}
\caption{ Experimental atom-atom radial distribution functions for water at
room temperature and pressure from reference~\cite{Soper2000}.  }
\label{correlation}
\end{figure}

@@

Numerous approaches have been developed to approximate the effect of water
as a solvent in atomistic calculations.  Each of these approaches gives
adequate description of some aspect of interactions with water, but none of
them is adequate for describing \emph{all} these interactions with an
accuracy close to that attained by \emph{ab initio} calculations.  The
theory of Lum, Chandler and Weeks (LCW)~\cite{LCW}, for instance, can
accurately describe the free energy cost of creating a cavity by placing a
solute in water, but does not lend itself to extensions treating the strong
interaction of water with hydrophilic solutes.  Treatment of water as a
continuum dielectric with a cavity surrounding each solute can give
accurate predictions of the energy of solvation of ions~\cite{latimer1939,
rashin1985, zhan1998, hsu1999, hildebrandt2004, hildebrandt2007}, but
provides no information about the size of this cavity.  In a physically
correct approach, the size of the cavity will naturally arise from a
balance between the free energy required to create the cavity, the
attraction between the water and the solute, and the steric repulsion which
opens up the cavity in the first place.

A very promising approach for an efficient continuum description of water
is that of classical density-functional theory (DFT), which is an approach
for evaluating the free energy and thermally averaged density of fluids in
an arbitrary external potential~\cite{ebner1976}.  Classical DFT is a
natural framework for creating a more flexible theory of hydrophobicity
that can describe interaction of water with arbitrary external
potentials---including potentials describing a \emph{strong} interaction
with a solute.

A number of exact properties are easily achieved in the density-functional
framework, such as the relationship between the pressure on a hard wall and
the contact density---which ensures a correct solvation free energy for
small hard solutes.  Much of the research on classical density-functional
theory has focussed on the hard-sphere fluid~\cite{curtin1985,
rosenfeld1989, rosenfeld1993, rosenfeld1997, tarazona1997, tarazona2000}.
This has led to a large number of very sophisticated functionals, such as
the fundamental-measure theory (FMT) functional~\cite{rosenfeld1989,
rosenfeld1993, rosenfeld1997, tarazona1997, tarazona2000}.  This functional
\needsworklater{has many nice properties}.

@@

\needsworklater{
Density-Functional Theory is most commonly used for calculating the
densities of electrons in quantum mechanical systems.  We take the
classical limit of this theory to construct a functional that will predict
the density of water molecules in a classical system.  The model used here
starts with a treatment of each individual water molecule as a purely
classical hard sphere.  The only constraint in this model is
the minimum distance between any two molecules is twice the hard sphere
radius.}

\needsworklater{ I still need to rewrite the entire introduction... }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

The simple yet complex water molecule is one of the most important 
substances for life and industry. The structure and geometry of the 
water molecule leads to a strong polarity and complex intermolecular
forces. The hydrogen bonds formed by water are very strong, creating
tetragonal clusters that lead to many of water's unusual properties,
such as the high melting, boiling, and critical temperatures as well as the
density maximum above the melting point.
The large dielectric constant is partially responsible for
water acting as a universal solvent, along with the ease of availability and
use.
There are many models used to describe water since so many different
processes occur in aqueous solution. It is challenging to describe all the 
relevant properties of water at once, especially if a low computational cost is
also desired.

Some models are primarily designed for use with homogeneous systems. Although
these 
models are useful in predicting some bulk properties, the interesting
chemistry that occurs in water cannot be described with a homogeneous-only
model. To
describe solvation and other processes in aqueous solution, a more robust theory
that 
can handle varying densities is needed.

Classical density-functional theory (DFT) provides an efficient way to evaluate
the free energy using the probability distribution of the fluid. The free energy 
is a functional
of the density function and is independent of any external
potentials\cite{ebner1976density}. The most common use of DFT is calculating ground
state electron densities in quantum mechanical systems. The Hohenberg-Kohn
theorem\cite{hohenberg1964inhomogeneous} allows us to write the ground state
energy in terms of a universal functional of the electron density and a separate
term including the external potential. Mermin\cite{mermin1965thermal} gives a
theorem extending that of Hohenberg and Kohn to nonzero temperatures:
\begin{equation}
  A(T) = \underset{n(r)}{min}\left\{ F[n(r),T] + \int V_{ext}(r) n(r)
d^3r\right\}
\end{equation}
This is the basis for classical DFT, where the relevant density is not the
electron density,
but the density of molecules. This can be used with a variety of 
external potentials, from the simple interaction with a 
hydrophobic wall to the more complex, stronger interaction with hydrophilic
solutes.

\subsection{Statistical Associating Fluid Theory}

The high hydrogen bonding and strong polar interactions in water characterize it
as an associating fluid, and thus it requires a different treatment than simple
fluids which are mainly defined by van der Waals forces and weak electrostatic
interations. The Statistical Associating Fluid Theory (SAFT) was designed for
use with fluids such as water, to account for the important association forces
(like hydrogen bonds) as well as any effects due to clusters or chains forming
in the liquid\cite{muller2001molecular}. It is based on Wertheim's first-order
thermodynamic perturbation theory (TPT1)
\cite{wertheim1984fluidsI,wertheim1984fluidsII,wertheim1986fluidsIII,
wertheim1986fluidsIV} which provides a useful way to account for strong
associative interations between molecules.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory and Methods}
We introduce a new classical density functional for water that
predicts the vapor pressure, vapor density, liquid density, bulk
surface tension, and enthalpy, as well as allows for liquid-vapor
coexistence.  The Helmholtz free energy functional is based on SAFT,
and is composed of the usual terms:
\begin{equation}
  F[n] = F_\textit{id}[n] + F_\textit{hs}[n]  +
F_\textit{disp}[n]+ F_\textit{assoc}[n],
\end{equation}
where $F_\textit{id}$ is the ideal gas free energy, $F_\textit{hs}$ is
the hard-sphere free energy, $F_\textit{disp}$ is the dispersion energy,
and $F_\textit{assoc}$ is the free energy of association.  In
addition, a chemical potential term is used, so we actually work in
the grand canonical ensemble.  In the following sections, we will
introduce the terms of this functional.

@@

\subsection{Other thermodynamic functions}

@@

The pressure would naturally be obtained by taking a partial
derivative with respect to volume at fixed temperature.  Since
\emph{volume} isn't a parameter in density-functional theory, we
consider the \emph{free energy per unit volume}:
\begin{align}
  F[n] &= f[n]V \\
  p[n] &= -\left(\frac{\partial F[n]}{\partial V}\right)_{T} \\
  &= -\left(\frac{\partial f[n]V}{\partial V}\right)_{T} \\
  &= -V\left(\frac{\partial f[n]}{\partial V}\right)_{T}
   - f[n]\left(\frac{\partial V}{\partial V}\right)_{T} \\
  &= n \left(\frac{\partial f[n]}{\partial n}\right)_{T} - f[n].
\end{align}

Figure~\ref{fig:pressure-with-isotherms} shows the calculated pressures for
isotherms ranging from room temperature to past the critical temperature. The
solid line is the theoretical coexistence curve calculated from the SAFT
functional. For a given isotherm, the pressure curve should intersect this
coexistence curve twice, at the vapor and liquid coexistence 
pressure corresponding to the vapor and liquid densities. Even
though the theoretical pressure isotherms drop below the coexistence curve,
this is non-physical and will not actually occur. The experimental data shows
how the actual density value will ``jump'' to the other side of 
the coexistence curve at that pressure value (for
temperatures less than the critical temperature), corresponding to a phase 
change in the material. For example, follow the room temperature isotherm
starting from high pressure and high density (top right of 
Figure~\ref{fig:pressure-with-isotherms}).
As the pressure decreases, the density decreases \emph{slightly} until reaching
the coexistence curve, at which point a phase change from liquid to vapor
will occur. The
functional matches well with the experimental data from NIST away from the
critical point. The slope of the pressure versus density curve is proportional
to the compressibility, which is close but slightly too large at room
temperature.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/equation-of-state}
\end{center}
\caption{The theoretical versus experimental phase diagram, vapor
  pressure versus temperature.  }
\label{fig:equation-of-state}
\end{figure}

The entropy is naturally obtained by taking a partial derivative with
respect to temperature:
\begin{align}
  S[n] &= \left(\frac{\partial F[n]}{\partial T}\right)_{V}
\end{align}
This derivative is trivial for the ideal gas and hard sphere free
energies, which are simply proportional to the temperature.  For the
association and dispersion free energies, it is a little more work,
but still not hard.

Once we have the entropy, finding the internal energy is simple:
\begin{align}
  U[n] &= F[n] + TS[n]
\end{align}
and the enthalpy is not much more work:
\begin{align}
  H[n] &= F[n] + TS[n] + pV.
\end{align}
The calculated heat of vaporization at atmospheric pressure is
$\Delta H_{vap}$= 39.41~kJ/mol at the normal boiling point of water,
373 K. This compares well to the experimental
value measured by NIST ($\Delta H_{vap}$= 40.65~kJ/mol\cite{nistwater}), with 
an error of only a few percent.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/temperature-versus-density}
\end{center}
\caption{The theoretical versus experimental coexistence curve. Away from
the critical temperature, the agreement is good. }
\label{fig:temperature-vs-density}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/finding-vapor-pressure}\\
\includegraphics[width=\columnwidth]{figs/near-critical-point}
\end{center}
\caption{By plotting the free energy versus density it is possible to
  determine a saturated liquid or vapor state by the common tangent
  method. The top plot illustrates the common tangent near room 
  temperature, while the bottom plot shows the same thing close to the 
  critical temperature. }
\label{fig:homogeneous}
\end{figure}

For reference, we plot the predicted vapor pressure (see
Figure~\ref{fig:equation-of-state}) and the coexistence curve (see
Figure~\ref{fig:temperature-vs-density}) against experimental values.  These
match those calculated by Clark \emph{et al}. The functional provides a good description of
the vapor pressure, which has an error of less than 
1\% compared with the experimental data. The coexistence curve in 
Figure~\ref{fig:temperature-vs-density} is similar to the pressure coexistence
curve in Figure~\ref{fig:pressure-with-isotherms} (solid line) but gives 
temperature coexistence values.
Near the critical temperature,
the agreement is off, and the theoretical prediction  of the temperature is quite a bit higher than
experiment. 
To find the densities for the coexistence curve, we use the
``common tangent'' rule. The two densities 
will be on the same tangent line to the free energy as illustrated in the top of Figure~\ref{fig:homogeneous}.
As the critical point approaches, the vapor and liquid coexistence densities get closer and closer together 
(see the bottom of Figure~\ref{fig:homogeneous}. This illustrates the difficulty
in correctly treating the critical temperature. As the vapor and liquid densities approach
each other, the free energy barrier between the two states gets smaller and it becomes
challenging to computationally find the correct minimum. This is the reason 
why the coexistence curves
in Figures~\ref{fig:pressure-with-isotherms}~and~\ref{fig:temperature-vs-density}
are not smooth around the critical point.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results and discussion}

\subsection{Water confined to a slit}

One of the most computationally simple ways to study confined water is by
constraining it to a slit where the water molecules can only move in one
dimension. Experimentally, this could represent two different situations.
The first, water between two perfectly aligned and smooth planes, is not common 
in experiment due to the difficulty of creating such an ideal situation. The
second
possibility is a large sphere brought close to a plane, which eliminates the 
need for perfect alignment.

Before looking at the liquid-solid one-dimensional slit, we set up a liquid-vapor
interface and plot the density profile.
The fluid density at a liquid-solid interface commonly displays
oscillatory behavior, but the presence of density oscillations at
liquid-vapor interfaces is more debated\cite{penfold2001structure}.
In fact, the \emph{meaning} of the density profile at a liquid-vapor
interface is clouded by the existence of long-wavelength perturbations
known as \emph{capillary waves}.  These waves are not accounted for in
density-functional theory (although the exact
density-functional theory would include them), with the result
that the density profile will be smoothed out by capillary wave effects. The liquid-vapor
interface is plotted in Figure~\ref{fig:liquid-vapor-profile} and shows
some small oscillations.

@@

To set up the calculations for the slit, we choose a chemical potential 
corresponding to a pressure of 1~atm and set the
temperature to a typical room temperature of 298~K. We apply 
a constraining potential outside the slit, where we choose a small non-zero
density for numerical reasons. The slits studied range in size from 0.1~nm to 
8.0~nm. We minimize the energy and calculate density, energy, and $X$ on a
lattice
with a spacing of approximately 0.01~nm (0.2~bohrs) between lattice points.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/density-1D}
\end{center}
\caption{ Density profile for a slab of water with molecules 
constrained to move in one dimension. The slit size is 4.0~nm and the dashed
line represents the saturated liquid density.}
\label{fig:density-1D}
\end{figure}

The density profile for a slit of size 4.0~nm is shown in Figure~\ref{fig:density-1D}. 
Near the interface, there are oscillations around the
liquid density which decrease near the center of the slit. We expect the center
of the slit to act approximately like bulk water, and the absence of
oscillations supports this. For larger slits (6.0~nm or more) the majority of
the water molecules in the slit act like the bulk.

Based on our computations, we determine the smallest slit that can contain 
liquid water. This result is
approximately 0.6~nm. For smaller slits, the initial liquid water inside will
vaporize. The results for larger liquid-filled slits are only metastable,
however, as a vapor-filled slit is thermodynamically favorable. Based on our
calculations for a pressure of 1~atm, the work needed to pull apart
the two hard walls comprising the slit is one to two orders of magnitude less
than the free energy of the liquid-filled slit. For the range of slit sizes we
studied, all the computational results of liquid-filled slits are
metastable.

Lum, Chandler and Weeks \cite{lum1999hydrophobicity} discuss the
stability of water between large hydrophobic plates and the critical separation
necessary for stability. For water at room temperature and atmospheric pressure,
plates with a separation smaller than about 100~nm will not contain stable liquid
water. They go on to say that confined water has a region of metastability down
to separations of about 5~nm. This is comparable to the slit separation where
our calculations show water inside the slit acting like the bulk liquid, which
we know to be a metastable solution. Forsman \emph{et al.}\cite{forsman1996computer}
also study water between two hard walls at constant chemical potential. They 
also conclude that the liquid state is likely metastable, and there is a free
energy barrier to reach the stable vapor phase.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/xassoc-1D}
\end{center}
\caption{The number of hydrogen bonds per molecule for a slab of water 
with molecules constrained to move in one dimension. Each molecule may
contribute to
four bonds, although each bond involves two molecules so the
actual number of bonds would be halved.}
\label{fig:xassoc-1D}
\end{figure}

The value of $X$ behaves as expected for water constrained to a slit. We
show in Figure~\ref{fig:xassoc-1D} the number of hydrogen bonds per water 
molecule, equal to $4(1-X)$, for a slit of 4.0~nm.
Outside the slit, there is almost no hydrogen bonding and inside the number of hydrogen
bonds is approximately equal to 3.5, which is expected for saturated liquid
water.

@@

We move into two dimensions by studying a single hydrophobic rod
immersed in water. Figure~\ref{fig:density-single-rod} shows density
profiles for three different radii, 0.3~nm, 0.5~nm, and 1~nm. There are the
expected oscillations outside the rod and as the distance away from the rod
increases
the density approaches that of the bulk saturated liquid density. The small
radius
rod has comparatively larger oscillations and a higher contact density.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/xassoc-single-rod}
\end{center}
\caption{ Number of bonds per molecule near the interface with a single
hydrophobic rod. There is a decrease in the hydrogen bonding near the surface 
of the rod, as expected at the interface of any hydrophobic wall. }
\label{fig:xassoc-single-rod}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/energy-vs-diameter}
\end{center}
\caption{ Free energy versus radius for a single hydrophobic rod
immersed in water. This should have an asymptote equal to the surface
tension at room temperature, and it agrees well with the surface tension in
Figure~\ref{fig:surface-tension}. }
\label{fig:energy-vs-diameter}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
{We develop a classical density functional theory for water by combining the
fundamental-measure theory functional--which describes hard sphere fluids--with
attractive interactions based on the Statistical Associating Fluid Theory
(SAFT-VR). Using this approximation with an additional length scale parameter, we are able to calculate the vapor pressure, saturated liquid density, surface tension, and enthalpy of water, as well as the number of hydrogen bonds and density oscillations formed near hydrophobic rods and spheres immersed in water. 
\needsworknow{For two rods being pulled apart, a transition occurs 
from a vapor state to a liquid state in the area between
the rods at approximately $(\pi-2)$ multipled by the radius. Even 
for nanoscale rods (radius 0.5~-~1.0~nm), we find that in the
vapor state, the force can be approximated
classically and is equal to 144~mN/m, or twice the bulk surface tension.
(Edit this and add spheres stuff)}
This functional is computationally efficient and provides a foundation for 
future work describing complex interactions in aqueous solution on an
atomic level.}
\tableofcontents
\end{abstract}
\maketitle

{\begin{itemize}
\color{blue}
\item Put finished things up here
\item[JRH] Rewrite abstract (copying from project)
\item[JRH] Cut the test-particle section
\item[JRH] Rewrite and join introductions
\item[JRH] Decide what to do with slit (cut)
\item[JRH] Combine and edit Other Thermo and Empirical Parameters sections
\color{red}
\item[JRH] Work on abstract
\item[DJR] Read and edit changes to abstract and introduction
\item[DJR] Look through theory section for stuff to cut
\item[DJR,JRH] Finalize figures in beginning
\item[JRH] Better/expanded description of ``macroscopic surface tension'' model in 2 rods section?
\item[JRH] Recalculate and clean up 2 rods figures and data - new code?
\item[DJR] Write text (or clean up existing) for hard sphere solutes
\item[EJK] Add 4-rods figures
\item[EJK] Add 4-rods text
\item[JRH,DJR] Rewrite conclusion
\end{itemize}
}

@@

Due to its large dielectric constant, small molecular size and strong
hydrogen-bonding network, water as ``the universal solvent'' is as
interesting as its anomalous thermodynamic properties.

@@

%\begin{figure}
%\includegraphics[width=6cm, angle=270]{figs/correlation}
%\caption{ Experimental atom-atom radial distribution functions for water at
%room temperature and pressure from reference~\cite{Soper2000}.\needsworknow{Either 
%mention this figure in the text or delete it.}  }
%\label{correlation}
%\end{figure}

Numerous approaches have been developed to approximate the effect of water
as a solvent in atomistic calculations.  Each of these approaches gives
adequate description of some aspect of interactions with water, but none of
them is adequate for describing \emph{all} these interactions with an
accuracy close to that attained by \emph{ab initio} calculations.  The
theory of Lum, Chandler and Weeks (LCW)~\cite{lum1999hydrophobicity}, for 
instance, can
accurately describe the free energy cost of creating a cavity by placing a
solute in water, but does not lend itself to extensions treating the strong
interaction of water with hydrophilic solutes.  Treatment of water as a
continuum dielectric with a cavity surrounding each solute can give
accurate predictions of the energy of solvation of ions~\cite{latimer1939,
rashin1985, zhan1998, hsu1999, hildebrandt2004, hildebrandt2007}, but
provides no information about the size of this cavity.  In a physically
correct approach, the size of the cavity will naturally arise from a
balance between the free energy required to create the cavity, the
attraction between the water and the solute, and the steric repulsion which
opens up the cavity in the first place.

A very promising approach for an efficient continuum description of water
is that of classical density-functional theory (DFT), which is an approach
for evaluating the free energy and thermally averaged density of fluids in
an arbitrary external potential~\cite{ebner1976density}. The Hohenberg-Kohn
theorem\cite{hohenberg1964inhomogeneous} allows us to write the ground state
energy in terms of a universal functional of the electron density and a separate
term including the external potential. Mermin\cite{mermin1965thermal} gives a
theorem extending that of Hohenberg and Kohn to nonzero temperatures:
\begin{equation}
  A(T) = \underset{n(r)}{min}\left\{ F[n(r),T] + \int V_{ext}(r) n(r)
d^3r\right\}.
\end{equation}
This is the basis for classical DFT, where the relevant density is not the
electron density, but the density of molecules. Classical DFT is a
natural framework for creating a more flexible theory of hydrophobicity
that can describe interaction of water with arbitrary external
potentials---including potentials describing a \emph{strong} interaction
with a solute.

A number of exact properties are easily achieved in the density-functional
framework, such as the relationship between the pressure on a hard wall and
the contact density---which ensures a correct solvation free energy for
small hard solutes.  Much of the research on classical density-functional
theory has focused on the hard-sphere fluid~\cite{curtin1985,
rosenfeld1989, rosenfeld1993, rosenfeld1997, tarazona1997, tarazona2000}.
This has led to a large number of very sophisticated functionals, such as
the fundamental-measure theory (FMT) functional~\cite{rosenfeld1989,
rosenfeld1993, rosenfeld1997, tarazona1997, tarazona2000}.  This functional
\needsworknow{has many nice properties}.

@@

\needsworknow{Add smooth transition from others' work to ours}

We use classical DFT to construct a new functional that will predict
the density of water molecules in a classical system.  The model we use
starts with a treatment of each individual water molecule as a purely
classical hard sphere and adds association terms based on Statistical 
Associating Fluid Theory (SAFT). Water, unlike simple
fluids which are mainly defined by van der Waals forces and weak electrostatic
interations, requires direct treatment of association effects. The strong 
hydrogen bonds and polar
interactions in water are ideal for describing with SAFT, which was designed for
use with such fluids. It is based on Wertheim's first-order
thermodynamic perturbation theory (TPT1)
\cite{wertheim1984fluidsI,wertheim1984fluidsII,wertheim1986fluidsIII,
wertheim1986fluidsIV} which provides a useful way to account for strong
associative interations between molecules. SAFT can account for the association energies
as well as any effects due to clusters or chains forming
in the liquid\cite{muller2001molecular}. \needsworknow{Expand or move this 
introduction to SAFT?}

\section{Theory and Methods}
We introduce a new classical density functional for water that
predicts the vapor pressure, vapor density, liquid density, bulk
surface tension, and enthalpy, as well as allows for liquid-vapor
coexistence.  The Helmholtz free energy functional is based on SAFT,
and is composed of the usual terms:
\begin{equation}
  F[n] = F_\textit{id}[n] + F_\textit{hs}[n]  +
F_\textit{disp}[n]+ F_\textit{assoc}[n],
\end{equation}
where $F_\textit{id}$ is the ideal gas free energy, $F_\textit{hs}$ is
the hard-sphere free energy, $F_\textit{disp}$ is the dispersion energy,
and $F_\textit{assoc}$ is the free energy of association.  In
addition, a chemical potential term is used, so we actually work in
the grand canonical ensemble.  In the following sections, we will
introduce the terms of this functional.

@@

@@

The pressure would naturally be obtained by taking a partial
derivative with respect to volume at fixed temperature. 
Figure~\ref{fig:pressure-with-isotherms} shows the calculated pressures for
isotherms ranging from room temperature to past the critical temperature. The
solid line is the theoretical coexistence curve calculated from the SAFT
functional. For a given isotherm, the pressure curve should intersect this
coexistence curve twice, at the vapor and liquid coexistence 
pressure corresponding to the vapor and liquid densities. Even
though the theoretical pressure isotherms drop below the coexistence curve,
this is non-physical and will not actually occur. The experimental data shows
how the actual density value will ``jump'' to the other side of 
the coexistence curve at that pressure value (for
temperatures less than the critical temperature), corresponding to a phase 
change in the material. The
functional matches well with the experimental data from NIST away from the
critical point. The slope of the pressure versus density curve is proportional
to the compressibility, which is close but slightly too large at room
temperature.

@@

%\begin{figure}
%\begin{center}
%\includegraphics[width=\columnwidth]{figs/equation-of-state}
%\end{center}
%\caption{The theoretical versus experimental phase diagram, vapor
%  pressure versus temperature.  }
%\label{fig:equation-of-state}
%\end{figure}

%The entropy is naturally obtained by taking a partial derivative with
%respect to temperature:
%\begin{align}
%  S[n] &= \left(\frac{\partial F[n]}{\partial T}\right)_{V}
%\end{align}
%This derivative is trivial for the ideal gas and hard sphere free
%energies, which are simply proportional to the temperature.  For the
%association and dispersion free energies, it is a little more work,
%but still not hard.

%Once we have the entropy, finding the internal energy is simple:
%\begin{align}
%  U[n] &= F[n] + TS[n]
%\end{align}
%and the enthalpy is not much more work:
%\begin{align}
%  H[n] &= F[n] + TS[n] + pV.
%\end{align}

The calculated heat of vaporization at atmospheric pressure is
$\Delta H_{vap}$= 39.41~kJ/mol at the normal boiling point of water,
373 K. This compares well to the experimental
value measured by NIST ($\Delta H_{vap}$= 40.65~kJ/mol\cite{nistwater}), with 
an error of only a few percent.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/temperature-versus-density}
\end{center}
\caption{The theoretical versus experimental coexistence curve. Away from
the critical temperature, the agreement is good.\needsworknow{(keep figure?
its the same as Clark paper)} }
\label{fig:temperature-vs-density}
\end{figure}

\subsection{Liquid-vapor interface}

We set up a liquid-vapor
interface and plot the density profile.
The fluid density at a liquid-solid interface commonly displays
oscillatory behavior, but the presence of density oscillations at
liquid-vapor interfaces is more debated\cite{penfold2001structure}.
In fact, the \emph{meaning} of the density profile at a liquid-vapor
interface is clouded by the existence of long-wavelength perturbations
known as \emph{capillary waves}.  These waves are not accounted for in
density-functional theory (although the exact
density-functional theory would include them), with the result
that the density profile will be smoothed out by capillary wave effects. 
The liquid-vapor
interface is plotted in Figure~\ref{fig:liquid-vapor-profile} and shows
some small oscillations.

@@

%\begin{figure}
%\begin{center}
%\includegraphics[width=\columnwidth]{figs/finding-vapor-pressure}\\
%\includegraphics[width=\columnwidth]{figs/near-critical-point}
%\end{center}
%\caption{By plotting the free energy versus density it is possible to
%  determine a saturated liquid or vapor state by the common tangent
%  method. The top plot illustrates the common tangent near room 
%  temperature, while the bottom plot shows the same thing close to the 
%  critical temperature. }
%\label{fig:homogeneous}
%\end{figure}

%For reference, we plot the predicted vapor pressure (see
%Figure~\ref{fig:equation-of-state}) and the coexistence curve (see
%Figure~\ref{fig:temperature-vs-density}) against experimental values.  These
%match those calculated by Clark \emph{et al}. 
%The functional provides a good description of
%the vapor pressure, which has an error of less than 
%1\% compared with the experimental data. The coexistence curve in 
%Figure~\ref{fig:temperature-vs-density} is similar to the pressure coexistence
%curve in Figure~\ref{fig:pressure-with-isotherms} (solid line) but gives 
%temperature coexistence values.
%Near the critical temperature,
%the agreement is off, and the theoretical prediction  of the temperature 
%is quite a bit higher than
%experiment. 
%To find the densities for the coexistence curve, we use the
%``common tangent'' rule. The two densities 
%will be on the same tangent line to the free energy as illustrated in the top of Figure~\ref{fig:homogeneous}.
%As the critical point approaches, the vapor and liquid coexistence densities get closer and closer together 
%(see the bottom of Figure~\ref{fig:homogeneous}. This illustrates the difficulty
%in correctly treating the critical temperature. As the vapor and liquid densities approach
%each other, the free energy barrier between the two states gets smaller and it becomes
%challenging to computationally find the correct minimum. This is the reason 
%why the coexistence curves
%in Figures~\ref{fig:pressure-with-isotherms}~and~\ref{fig:temperature-vs-density}
%are not smooth around the critical point.
% 
% To set up the calculations for the slit, we choose a chemical potential 
% corresponding to a pressure of 1~atm and set the
% temperature to a typical room temperature of 298~K. We apply 
% a constraining potential outside the slit, where we choose a small non-zero
% density for numerical reasons. The slits studied range in size from 0.1~nm to 
% 8.0~nm. We minimize the energy and calculate density, energy, and $X$ on a
% lattice
% with a spacing of approximately 0.01~nm (0.2~bohrs) between lattice points.
% 
% \begin{figure}
% \begin{center}
% \includegraphics[width=\columnwidth]{figs/density-1D}
% \end{center}
% \caption{ Density profile for a slab of water with molecules 
% constrained to move in one dimension. The slit size is 4.0~nm and the dashed
% line represents the saturated liquid density.}
% \label{fig:density-1D}
% \end{figure}
% 
% The density profile for a slit of size 4.0~nm is shown in Figure~\ref{fig:density-1D}. 
% Near the interface, there are oscillations around the
% liquid density which decrease near the center of the slit. We expect the center
% of the slit to act approximately like bulk water, and the absence of
% oscillations supports this. For larger slits (6.0~nm or more) the majority of
% the water molecules in the slit act like the bulk. 
% 
% Based on our computations, we determine the smallest slit that can contain 
% liquid water. This result is
% approximately 0.6~nm. For smaller slits, the initial liquid water inside will
% vaporize. The results for larger liquid-filled slits are only metastable,
% however, as a vapor-filled slit is thermodynamically favorable. Based on our
% calculations for a pressure of 1~atm, the work needed to pull apart
% the two hard walls comprising the slit is one to two orders of magnitude less
% than the free energy of the liquid-filled slit. For the range of slit sizes we
% studied, all the computational results of liquid-filled slits are
% metastable. 
% 
% Lum, Chandler and Weeks \cite{lum1999hydrophobicity} discuss the
% stability of water between large hydrophobic plates and the critical separation
% necessary for stability. For water at room temperature and atmospheric pressure,
% plates with a separation smaller than about 100~nm will not contain stable liquid
% water. They go on to say that confined water has a region of metastability down
% to separations of about 5~nm. This is comparable to the slit separation where
% our calculations show water inside the slit acting like the bulk liquid, which
% we know to be a metastable solution. Forsman \emph{et al.}\cite{forsman1996computer}
% also study water between two hard walls at constant chemical potential. They 
% also conclude that the liquid state is likely metastable, and there is a free
% energy barrier to reach the stable vapor phase. 
% 
% \begin{figure}
% \begin{center}
% \includegraphics[width=\columnwidth]{figs/xassoc-1D}
% \end{center}
% \caption{The number of hydrogen bonds per molecule for a slab of water 
% with molecules constrained to move in one dimension. Each molecule may
% contribute to
% four bonds, although each bond involves two molecules so the
% actual number of bonds would be halved.}
% \label{fig:xassoc-1D}
% \end{figure}
% 
% The value of $X$ behaves as expected for water constrained to a slit. We
% show in Figure~\ref{fig:xassoc-1D} the number of hydrogen bonds per water 
% molecule, equal to $4(1-X)$, for a slit of 4.0~nm.
% Outside the slit, there is almost no hydrogen bonding and inside the number of hydrogen
% bonds is approximately equal to 3.5, which is expected for saturated liquid
% water.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results and discussion}

@@

We move into two dimensions by studying a single hydrophobic rod
immersed in water. We use a lattice spacing of about 0.01~nm (0.2 bohrs), 
which is a reasonable resolution based on several lattice spacing calculations.
The results are within 0.5\% of the free energy and a few percent of the contact
density for the highest resolution tested.
Figure~\ref{fig:density-single-rod} shows density
profiles for three different radii, 0.3~nm, 0.5~nm, and 1~nm. There are the
expected oscillations outside the rod and as the distance away from the rod
increases
the density approaches that of the bulk saturated liquid density. The small
radius
rod has comparatively larger oscillations and a higher contact density.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/energy-vs-diameter}
\end{center}
\caption{ Free energy versus radius for a single hydrophobic rod
immersed in water. This should have an asymptote equal to the surface
tension at room temperature, and it agrees well with the surface tension in
Figure~\ref{fig:surface-tension}.\needsworknow{Add more data at small radius 
and include an inset.} }
\label{fig:energy-vs-diameter}
\end{figure}
